---
title: "GBBO Data Cleaning/Joining"
subtitle: "v1"
author: "Konnor Herbst"
format:
  pdf:
    fontsize: 11pt
    toc: true
    toc-depth: 7
    colorlinks: true
    linestretch: 1
    fig-width: 12
    fig-height: 8
    fig-align: center
    df-print: kable
execute:
  cache: FALSE
  echo: FALSE
  warning: false
  message: false
output: pdf_document
---

\newpage

```{r, echo = FALSE}
# helpers
set.seed(489)
`%notin%` = Negate(`%in%`)

options(digits = 3)

# packages
library(forcats)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(janitor)
library(tibble)
```

# Data Cleaning

I combine and clean several sources of data on the Netflix seasons of
\textit{The Great British Bake Off}, including:

\begin{itemize}
  \item Judging data prepared by \href{https://github.com/nathangiusti/BakeOff}{Nathan Giusti}.
  \item Baker demogrpahic (age/occupation) data from the
  \href{https://cran.r-project.org/web/packages/bakeoff/index.html}{R \texttt{bakeoff} package} and \href{https://www.kaggle.com/datasets/sarahvitvitskiy/great-british-bake-off-results-seasons-1-14}{Kaggle}
  \item Hand-collected 7-day UK viewer counts from
  \href{https://www.barb.co.uk/viewing-data/most-viewed-programmes/}{BARB},
  used to supplement missing viewer data for later seasons.
\end{itemize}

## Giusti's Data

The technical challenge score is different from the signature and showstopper
scores because it is a ranking rather than an absolute component score. This
creates a scaling issue because the number of contestants changes across rounds
as bakers are eliminated. For example, first place out of twelve bakers carries
more information than first place out of three bakers, and third place out of
three is the worst performance in that episode even though rank three would be
strong early in the season.

For example,

\begin{itemize}
  \item Round 1: rank 1 out of 12 is very strong.
  \item Round 9: rank 1 out of 3 is still strong, but less informative.
  \item Round 1: rank 6 out of 12 is middle.
  \item Round 9: rank 3 out of 3 is worst.
\end{itemize}

For this reason, I make another variable so that we need not treat the raw technical rank as directly comparable
across rounds. I transform it into a within-episode relative rank,

$$
\operatorname{tech\_scaled}_{ie}
=
1 - 2\frac{\operatorname{tech}_{ie} - 1}{n_e - 1}.
$$

where $n_e$ is the number of bakers remaining in episode $e$. This maps the
technical result to a stable relative scale: $1$ is best, $0$ is middle of the
field, and $-1$ is worst.

To keep the model predictors on a common scale, I also rescale this to mathc the range of the signature and
showstopper totals. Each signature and showstopper score combines flavor,
appearance, and bake components, so the raw totals range from $-3$ to $3$. I then multiply the above by $3$ to match

After this signature, showstopper, and technical performance are
all coded on the same $[-3,3]$ scale. Positive values indicate stronger
performance, negative values indicate weaker performance, and zero corresponds
to a neutral or middle performance.

```{r giusti-data, echo=FALSE}
giusti_file <- "human_judging_giusti.csv"

judging_raw <- read_csv(giusti_file, show_col_types = FALSE)

judging <- judging_raw %>%
  clean_names() %>%
  rename(
    star_baker = winner,
    tech_rank  = tech
  ) %>%
  mutate(
    episode = interaction(series, round, drop = TRUE)
  ) %>%
  mutate(
    across(
      .cols = matches("handshake|eliminated|star_baker") & where(is.numeric),
      .fns  = ~ replace_na(.x, 0)
    ),
    eliminated = case_when(
      round == 10 & star_baker == 1 ~ 0,
      round == 10 & star_baker == 0 ~ 1,
      TRUE ~ eliminated
    )
  ) %>%
  select(-contains("_handshake")) %>%
  mutate(
    sig_sum = rowSums(
      across(c(signature_bake, signature_flavor, signature_looks)),
      na.rm = TRUE
    ),
    showstopper_sum = rowSums(
      across(c(showstopper_bake, showstopper_flavor, showstopper_looks)),
      na.rm = TRUE
    ),
    sig_ss_sum = sig_sum + showstopper_sum
  ) %>%
  group_by(series, round) %>%
  mutate(
    n_remaining = n(),
    tech_rank_scaled = if_else(
      n_remaining > 1,
      1 - 2 * (tech_rank - 1) / (n_remaining - 1),
      NA_real_
    ),
    tech_rank_scaled = 3 * tech_rank_scaled
  ) %>%
  ungroup() %>%
  group_by(contestant, series) %>%
  arrange(round, .by_group = TRUE) %>%
  mutate(
    series_winner = as.integer(last(round) == 10 & last(star_baker) == 1),
    eliminated_round = last(round)
  ) %>%
  ungroup() %>%
  mutate(
    baker_id = interaction(series, contestant, drop = TRUE)
  ) %>%
  arrange(series, round, contestant) %>%
  select(
    contestant,
    series,
    round,
    episode,
    tech_rank_scaled,
    sig_sum,
    showstopper_sum,
    tech_rank,
    n_remaining,
    series_winner,
    eliminated_round,
    baker_id,
    everything()
  )

judging_sb <- judging

judging_elim <- judging %>%
  filter(round <= 9) %>%
  group_by(episode) %>%
  filter(sum(eliminated == 1, na.rm = TRUE) == 1) %>%
  ungroup()
```

## Kaggle

```{r kaggle-data}
kaggle_dir <- "kaggleVitvitskiy"

winners <- read_csv(
  file.path(kaggle_dir, "GBBO_Winners.csv"),
  show_col_types = FALSE
)

gbbo_kaggle <- map_dfr(1:14, function(s) {
  file <- file.path(kaggle_dir, paste0("Season_", s, ".csv"))

  read_csv(
    file,
    col_types = cols(.default = col_character()),
    show_col_types = FALSE
  ) %>%
    mutate(season = s, .before = 1)
})

gbbo_kaggle_long <- gbbo_kaggle %>%
  pivot_longer(
    cols = matches("^Round_([1-9]|10)_"),
    names_to = c("round", "variable"),
    names_pattern = "^Round_([1-9]|10)_(.*)$",
    values_to = "value"
  ) %>%
  mutate(
    round = as.integer(round),
    variable = str_to_lower(variable),
    variable = str_replace_all(variable, "\\s+", "_")
  ) %>%
  pivot_wider(
    names_from = variable,
    values_from = value
  ) %>%
  clean_names() %>%
  mutate(series = season - 3) %>%
  filter(series >= 5) %>%
  rename(contestant = baker) %>%
  select(
    -technicals,
    -handshakes,
    -star_bakers,
    -handshake,
    -contains("_")
  ) %>%
  drop_na() %>%
  arrange(series, round, contestant)

write_csv(gbbo_kaggle, file.path(kaggle_dir, "gbbo_kaggle_seasons_1_14.csv"))
write_csv(gbbo_kaggle_long, file.path(kaggle_dir, "kaggle_long_clean.csv"))
```

to join with giustis data we can simply omit those seasons that are not included in netflx,

```{r join-giusti-kaggle}
giusti_kaggle_cln <- judging %>%
  left_join(
    gbbo_kaggle_long,
    by = c("series", "round", "contestant")
  ) %>%
  clean_names() %>%
  select(-any_of("technical"))

write_csv(giusti_kaggle_cln, "giusti_kaggle_cln.csv")
```

here thankfully we see mostly agreement on the technical (did giusti use this data to get it? I am not certain) with discrepency only on 4 judgings - I keep giustis here

## R bakeoff Package

```{r rbakeoff-package}
rbakeoff_dir <- "RbakeoffPackage"

read_rbakeoff_csv <- function(filename) {
  read_csv(
    file.path(rbakeoff_dir, filename),
    show_col_types = FALSE
  ) %>%
    clean_names() %>%
    mutate(series = series - 3) %>%
    filter(series >= 5)
}

# for another time maybe flavor or whatever text analysis
bakes_raw <- read_rbakeoff_csv("bakes_raw.csv")
challenges <- read_rbakeoff_csv("challenges.csv")
ratings_raw <- read_rbakeoff_csv("ratings_raw.csv")

ratings <- read_rbakeoff_csv("ratings.csv") %>%
  rename(round = episode)

giusti_kaggle_ratings_cln <- giusti_kaggle_cln %>%
  left_join(
    ratings,
    by = join_by(series, round)
  ) %>%
  select(
    -any_of(c(
      "bbc_iplayer_requests",
      "episode_count",
      "us_season",
      "us_airdate",
      "season",
      "network_rank",
      "viewers_28day"
    ))
  ) %>%
  select(
    uk_airdate,
    episode,
    series,
    round,
    viewers_7day,
    channels_rank,
    baker_id,
    contestant,
    age,
    gender,
    hometown,
    occupation,
    everything()
  )

write_csv(giusti_kaggle_ratings_cln, "cur_cleanest_version.csv")
```

the viewer data is unavailable in this data from series 8 onwards, perhaps IO can fill that in by hand? I went to the British broadcasting groups data BARB and got that,

# Data Gathering Myself

## BARB viewer data - hand entry

For the weeks when GBBO aired in the UK I compiled the 7 day viewing data/counts from ![BARB](https://www.barb.co.uk/viewing-data/most-viewed-programmes/)

My numbers are for post 2021/netflix seasons to add to what was in the bakeoff package, the bakeoff package seems to have gotten its data from BARB as well

```{r barb-manual-entry}
gbbo_uk_airdates <- tribble(
  ~series, ~round, ~uk_airdate, ~viewers_7day, ~channels_rank,
  8, 1, "2020-09-22", 11211907, 1,
  8, 2, "2020-09-29", 10806516, 1,
  8, 3, "2020-10-06", 10698161, 1,
  8, 4, "2020-10-13", 10629433, 1,
  8, 5, "2020-10-20", 10657511, 2,
  8, 6, "2020-10-27", 10370564, 3,
  8, 7, "2020-11-03", 10866618, 2,
  8, 8, "2020-11-10", 10911563, 3,
  8, 9, "2020-11-17", 10827776, 8,
  8, 10, "2020-11-24", 11744692, 1,
  9, 1, "2021-09-21", 9610000, 3,
  9, 2, "2021-09-28", 8978000, 2,
  9, 3, "2021-10-05", 8694000, 2,
  9, 4, "2021-10-12", 8539000, 2,
  9, 5, "2021-10-19", 8711000, 3,
  9, 6, "2021-10-26", 9151000, 2,
  9, 7, "2021-11-02", 9084000, 2,
  9, 8, "2021-11-09", 9051000, 3,
  9, 9, "2021-11-16", 9385000, 3,
  9, 10, "2021-11-23", 9607000, 3,
  10, 1, "2022-09-13", 8301000, 1,
  10, 2, "2022-09-20", 7601000, 5,
  10, 3, "2022-09-27", 7857000, 3,
  10, 4, "2022-10-04", 7770000, 3,
  10, 5, "2022-10-11", 8317000, 2,
  10, 6, "2022-10-18", 7794000, 3,
  10, 7, "2022-10-25", 7512000, 3,
  10, 8, "2022-11-01", 7414000, 4,
  10, 9, "2022-11-08", 7699000, 10,
  10, 10, "2022-11-15", 8125000, 10,
  11, 1, "2023-09-26", 7836000, 2,
  11, 2, "2023-10-03", 7543000, 2,
  11, 3, "2023-10-10", 7282000, 3,
  11, 4, "2023-10-18", 7120000, 4,
  11, 5, "2023-10-24", 5885000, 4,
  11, 6, "2023-10-31", 6320000, 3,
  11, 7, "2023-11-07", 6937000, 3,
  11, 8, "2023-11-14", 6771000, 4,
  11, 9, "2023-11-21", 6473000, 11,
  11, 10, "2023-11-28", 6750000, 10,
  12, 1, "2024-09-24", 6621000, 3,
  12, 2, "2024-10-01", 6520000, 3,
  12, 3, "2024-10-08", 6497000, 3,
  12, 4, "2024-10-15", 6409000, 4,
  12, 5, "2024-10-22", 6316000, 4,
  12, 6, "2024-10-29", 6139000, 4,
  12, 7, "2024-11-05", 6432000, 3,
  12, 8, "2024-11-12", 6386000, 4,
  12, 9, "2024-11-19", 6260000, 10,
  12, 10, "2024-11-26", 6743000, 10
) %>%
  mutate(
    uk_airdate = as.Date(uk_airdate)
  )

dir.create("data/manual", recursive = TRUE, showWarnings = FALSE)

write_csv(
  gbbo_uk_airdates,
  "data/manual/gbbo_uk_airdates_viewers.csv"
)
```

then combine with the other data

```{r combine-viewer-data}
other_dates <- giusti_kaggle_ratings_cln %>%
  select(uk_airdate, series, round, viewers_7day, channels_rank) %>%
  mutate(
    uk_airdate = as.Date(uk_airdate),
    viewers_7day = viewers_7day * 1000000
  )

dates <- gbbo_uk_airdates %>%
  bind_rows(other_dates) %>%
  distinct(series, round, .keep_all = TRUE) %>%
  arrange(series, round)

full_cln_dat <- giusti_kaggle_ratings_cln %>%
  select(-any_of(c("uk_airdate", "viewers_7day", "channels_rank"))) %>%
  left_join(dates, by = c("series", "round")) %>%
  select(
    uk_airdate,
    episode,
    series,
    round,
    viewers_7day,
    channels_rank,
    baker_id,
    contestant,
    age,
    gender,
    hometown,
    occupation,
    everything()
  ) %>%
  arrange(series, round, contestant)

write_csv(full_cln_dat, "full_cln_gbbo.csv")
```
